---
title: "Conclusions"
output: html_document
date: "2026-05-04"
editor_options:
  markdown:
    wrap: 72
---

## Setup

We start from the Phase 1 cleaned dataset (`distress_primary.csv`),
not the raw panel. That handles outcome-side leakage (RPY_3YR_RT,
earnings columns dropped upstream), the COVID-pause years (filtered to
year <= 2019), and the missing-coverage drops in one place so this
script can stay focused on modeling.

```{r setup, message = FALSE, warning = FALSE}
library(tidymodels)
library(tidyverse)
library(here)
library(lme4)
library(performance)
library(glmnet)

panel <- read_csv(here("data", "processed", "distress_primary.csv"),
                  show_col_types = FALSE)

# CSV round-trip strips factor levels -- restore them so the recipe
# treats categoricals as categorical, not as text.
panel <- panel %>%
  mutate(across(c(CONTROL, PREDDEG, REGION, LOCALE), as.factor)) %>%
  # `distressed` is derived from CDR3 (CDR3 > 0.10), so including it
  # as a predictor would be perfect leakage.
  select(-any_of("distressed"))
```

## Grouped train/test split

Schools appear in multiple years. A random row-level split puts the
same UNITID in both train and test, which inflates R^2 because the
model just learns school identity. `group_initial_split` keeps every
school on one side of the split.

```{r split}
set.seed(42)

panel_split <- group_initial_split(panel, group = UNITID, prop = 0.8)
train_data  <- training(panel_split)
test_data   <- testing(panel_split)

c(train_schools = n_distinct(train_data$UNITID),
  test_schools  = n_distinct(test_data$UNITID),
  train_rows    = nrow(train_data),
  test_rows     = nrow(test_data))
```

## Recipe

Median imputation is fit on the training data and then applied to test
(no test-set leakage). NA factor levels become an explicit `unknown`
level so we don't drop rows. Predictors are normalized so the L1/L2
penalties act on the same scale.

```{r recipe}
rec <- recipe(CDR3 ~ ., data = train_data) %>%
  update_role(UNITID, INSTNM, OPEID, year, new_role = "id") %>%
  step_unknown(all_nominal_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

## Tune penalty and mixture together

5-fold grouped CV on the training set, tuning both `penalty` (lambda)
and `mixture` (alpha) so the model can sit anywhere on the LASSO-Ridge
continuum instead of being pinned at alpha = 0.5.

```{r tune, message = FALSE, warning = FALSE}
en_spec <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet")

wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(en_spec)

set.seed(42)
folds <- group_vfold_cv(train_data, group = UNITID, v = 5)

en_grid <- grid_regular(
  penalty(range = c(-4, 0)),
  mixture(range = c(0, 1)),
  levels = c(penalty = 20, mixture = 6)
)

tuned <- tune_grid(
  wf,
  resamples = folds,
  grid      = en_grid,
  metrics   = metric_set(yardstick::rmse, yardstick::rsq)
)

show_best(tuned, metric = "rmse", n = 5)
```

## Fit final model on full train, evaluate on test

```{r fit}
best <- select_best(tuned, metric = "rmse")
best

final_wf  <- finalize_workflow(wf, best)
final_fit <- fit(final_wf, data = train_data)

test_preds <- predict(final_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(CDR3))

en_metrics <- test_preds %>%
  metrics(truth = CDR3, estimate = .pred)

rmse_en <- en_metrics %>% filter(.metric == "rmse") %>% pull(.estimate)
r2_en   <- en_metrics %>% filter(.metric == "rsq")  %>% pull(.estimate)

# Null model = predict the training mean for every test row.
mean_train <- mean(train_data$CDR3, na.rm = TRUE)
rmse_null  <- sqrt(mean((test_data$CDR3 - mean_train)^2, na.rm = TRUE))

tibble(
  model = c("Null (train mean)", "Elastic Net (tuned, grouped CV)"),
  rmse  = c(rmse_null, rmse_en)
)

reduction_pct <- (rmse_null - rmse_en) / rmse_null * 100
```

## Coefficients

The interpretive payoff of the elastic net -- which institutional
features actually carry weight after penalization. Nonzero coefficients
only, sorted by magnitude.

```{r coefs}
en_coefs <- final_fit %>%
  extract_fit_parsnip() %>%
  tidy() %>%
  filter(term != "(Intercept)", estimate != 0) %>%
  mutate(abs_est = abs(estimate)) %>%
  arrange(desc(abs_est)) %>%
  select(term, estimate, penalty)

en_coefs %>% print(n = 25)

n_nonzero <- nrow(en_coefs)
```

## Mixed-effects variance decomposition

Updated ICC with both school and year as random effects. The original
EDA fit only `(1 | UNITID)`; adding `(1 | year)` lets us separate
school-level variance from year-to-year drift.

```{r icc, message = FALSE, warning = FALSE}
icc_model <- lmer(CDR3 ~ 1 + (1 | UNITID) + (1 | year), data = panel)
summary(icc_model)
icc(icc_model, by_group = TRUE)

icc_vals <- icc(icc_model, by_group = TRUE)
icc_school <- icc_vals %>% filter(Group == "UNITID") %>% pull(ICC)
icc_year   <- icc_vals %>% filter(Group == "year")   %>% pull(ICC)
```

## Final ideas

#### One

-   **`r round(icc_school * 100, 1)`%** of the variance in default rates
    sits between schools (school-level ICC), and only
    **`r round(icc_year * 100, 1)`%** is year-to-year drift.

If I pick two schools at random, most of the difference in default
rates comes from *which school they are* -- not random fluctuation
within a school over time, and not the cohort year either.

#### Two

-   Null RMSE: **`r round(rmse_null, 4)`**
-   Elastic Net RMSE (tuned, grouped CV): **`r round(rmse_en, 4)`**
-   Test R^2: **`r round(r2_en, 3)`**
-   That's a **`r round(reduction_pct, 1)`%** reduction in RMSE over
    the null.
-   Best hyperparameters: penalty = `r signif(best$penalty, 3)`,
    mixture = `r best$mixture`. The model retained
    **`r n_nonzero`** nonzero coefficients.

#### Final

-   Institution features reduce prediction error meaningfully relative
    to the null, even after blocking school identity out of the test
    set (so the gain is *generalization* across schools, not memorizing
    them).
-   Most of the variation in CDR3 is persistent at the school level
    (`r round(icc_school * 100, 1)`% ICC), which means the elastic net
    is recovering structural school traits -- not transient cohort
    effects.
-   Compared with the earlier random-split run (R^2 ~ 0.33, RMSE
    reduction ~ 18%), the grouped split is the honest number. Random
    splits on panel data overstate performance because the test set
    shares schools with the training set.
