---
title: "Conclusions"
output: html_document
date: "2026-05-04"
editor_options: 
  markdown: 
    wrap: 72
---

## Elastic Net

```{r}
library(tidymodels)
library(rsample)
library(tidyverse)
```

## Split Data

Need the matrices to align in size, be careful

```{r}
library(glmnet)
library(dplyr)

set.seed(42)
train_index <- sample(seq_len(nrow(panel)), size = 0.8 * nrow(panel))

train_data <- panel[train_index, ]
test_data  <- panel[-train_index, ]

train_clean <- train_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

test_clean <- test_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
```

```{r}
#dim(X_test)
#length(X_train)
```

```{r}
X_train <- model.matrix(CDR3 ~ . - UNITID - INSTNM - OPEID, data = train_clean)[, -1]
y_train <- train_clean$CDR3

X_test  <- model.matrix(CDR3 ~ . - UNITID - INSTNM - OPEID, data = test_clean)[, -1]
y_test  <- test_clean$CDR3
```

## Elastic Net Model

Give it a second it does need a minute to run!

```{r}
cv_model <- cv.glmnet(X_train, y_train, alpha = 0.5)

# PREDICTION 

preds <- predict(cv_model, s = "lambda.min", newx = X_test)

```

## Stronger Evaluation Possibly

```{r}
rmse <- sqrt(mean((y_test - preds)^2))
rmse
```

```{r}
mean_pred <- mean(y_train)
rmse_null <- sqrt(mean((y_test - mean_pred)^2))
rmse_null
```

## First Comparison (Baseline and Elastic Model)

Using institutional features alone, we can predict default rates with
\~4.4% error, explaining \~X% of variation.”

```{r}
ss_res <- sum((y_test - preds)^2)
ss_tot <- sum((y_test - mean(y_test))^2)

r2 <- 1 - (ss_res / ss_tot)
r2
```

## Addition to original EDA Lmer Model

```{r}
icc_model <- lmer(CDR3 ~ 1 + (1 | UNITID) + (1 | YEAR), data = panel)

```

Try to use year as something to explain the last of the variance % with
the Elastic model.

## Final Ideas

#### One

-   72% of the variance in default rates is between schools \*(from
    original EDA report (assignment 2))

This is important becasue: If I randomly pick two schools, most of the
difference in default rates is due to which school they are — not random
fluctuation within a school over time.

#### Two

-   The Elastic Net reduced error from: 0.0789 → 0.0438
-   Baseline compared to EN
-   That’s \~44% reduction in RMSE

#### Final

-   Institution features reduce prediction error substantially (almost
    44%), indicating extremely meaningful explanatory power.
-   Additionally, most of the variation ( almost 72%) exists at the
    school level, suggesting strong persistent institutional
    differences.
