---
title: "distress_EDA"
output: pdf_document
---

```{r}
library(here)
library(ggplot2)
library(dplyr)
library(readr)

panel <- read_csv(here("data", "processed", "distress_primary.csv"),
                   show_col_types = FALSE)
```



1. **Distribution of `CDR3`** — histogram, by year. Does the distribution shift over time? (If yes, year must be a predictor or the data must be split by year.)


```{r}
ggplot(panel, aes(x = CDR3)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") + 
  facet_wrap(~ year)+ 
  labs(title = "Distribution of Three Year Default Rate by Year", x = "CDR3", y = "Count") +
  theme_minimal()

```
In 2022 and 2023, 100% of all schools have CDR3 = 0. This is because after covid there was a freeze on student loan repayments so no one could default. This is justification for these years to be dropped from the analysis, as they contain no signal on distress. 

The distributions of 3 year default rate does not appear to shift much over time. 

   
2. **Within-school vs between-school variance in `CDR3`.**

```{r}
by_school <- panel %>%
    group_by(UNITID) %>%
    summarise(mean_cdr3 = mean(CDR3), sd_cdr3 = sd(CDR3), n = n(),
              .groups = "drop")

var_between <- var(by_school$mean_cdr3)
var_within  <- mean(by_school$sd_cdr3^2, na.rm = TRUE)
var_total   <- var(panel$CDR3)

c(between = var_between / var_total,
  within  = var_within  / var_total)
```

**Intraclass correlation (ICC)** from a random-intercept model -- the canonical "proportion of total variance attributable to school identity."

```{r}
library(lme4)
library(performance)

icc_model <- lmer(CDR3 ~ 1 + (1 | UNITID), data = panel)
icc(icc_model)
```

**Spaghetti plot** -- every line is one school's CDR3 trajectory. Sample of 75 schools per sector for readability.

```{r}
set.seed(42)
sample_ids <- panel %>%
  distinct(UNITID, CONTROL) %>%
  group_by(CONTROL) %>%
  slice_sample(n = 75) %>%
  pull(UNITID)

panel %>%
  filter(UNITID %in% sample_ids) %>%
  ggplot(aes(x = year, y = CDR3, group = UNITID, color = CONTROL)) +
  geom_line(alpha = 0.35) +
  facet_wrap(~ CONTROL) +
  labs(title = "CDR3 trajectories by sector (sample of 75 schools per sector)",
       x = "Cohort year", y = "3-year default rate") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Mean vs within-school SD** -- are high-distress schools also more volatile year to year?

```{r, warning = FALSE}
by_school_ctrl <- panel %>%
  group_by(UNITID) %>%
  summarise(
    mean_cdr3 = mean(CDR3),
    sd_cdr3   = sd(CDR3),
    CONTROL   = first(CONTROL),
    .groups   = "drop"
  )

# Log scale drops zeros and NAs silently -- report how many points we lose.
n_zero_sd <- sum(by_school_ctrl$sd_cdr3 == 0, na.rm = TRUE)
n_na_sd   <- sum(is.na(by_school_ctrl$sd_cdr3))
message(sprintf(
  "%d schools with zero SD, %d with NA SD (will be dropped on log scale)",
  n_zero_sd, n_na_sd
))

ggplot(by_school_ctrl, aes(x = mean_cdr3, y = sd_cdr3, color = CONTROL)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "School-level mean vs within-school SD of CDR3",
       subtitle = "Log-log scale; black line is OLS fit",
       x = "Mean CDR3 (log scale)",
       y = "Within-school SD of CDR3 (log scale)",
       color = "Sector") +
  theme_minimal()
```
295 schools with zero SD, 0 with NA SD (will be dropped on log scale)



3. **Univariate plots** of `CDR3` against the top candidate predictors: `CONTROL`, `PREDDEG`, `PCTPELL`, `ADM_RATE`, `C150`, `GRAD_DEBT_MDN`, `NPT4`

Categorical predictors (`CONTROL`, `PREDDEG`) -- CDR3 distribution by level.

```{r, warning = FALSE}
panel %>%
  select(CDR3, CONTROL, PREDDEG) %>%
  tidyr::pivot_longer(-CDR3, names_to = "variable", values_to = "level") %>%
  ggplot(aes(x = level, y = CDR3)) +
  geom_boxplot(outlier.alpha = 0.15, fill = "steelblue", alpha = 0.5) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "CDR3 by categorical predictor",
       x = NULL, y = "CDR3") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Continuous predictors -- scatter with a GAM smooth. `alpha = 0.1` handles overplotting; GAM scales better than LOESS at this sample size. NAs are filtered per-predictor so every facet has clean data.

```{r, warning = FALSE, message = FALSE}
panel %>%
  select(CDR3, PCTPELL, ADM_RATE, C150, GRAD_DEBT_MDN, NPT4) %>%
  tidyr::pivot_longer(-CDR3, names_to = "variable", values_to = "value") %>%
  filter(!is.na(value), !is.na(CDR3)) %>%
  ggplot(aes(x = value, y = CDR3)) +
  geom_point(alpha = 0.1, size = 0.6, na.rm = TRUE) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
              se = FALSE, color = "firebrick", linewidth = 0.7,
              na.rm = TRUE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "CDR3 vs continuous predictors",
       subtitle = "Red line is GAM smooth (cubic spline)",
       x = NULL, y = "CDR3") +
  theme_minimal()
```

   
   
   
   
4. **Bivariate correlation heatmap** of the continuous features — flag any pairs with |r| > 0.8 for the penalized-regression stage.

```{r, fig.width = 12, fig.height = 10, warning = FALSE}
# All numeric predictor columns. Exclude identifiers, time, and the outcome
# (CDR3) -- feature-feature relationships are what matter here for
# multicollinearity diagnostics going into LASSO.
numeric_predictors <- panel %>%
  select(where(is.numeric)) %>%
  select(-any_of(c("UNITID", "OPEID", "year", "CDR3")))

# Drop columns that are entirely NA or have zero variance -- these produce
# NA correlations and break hclust downstream.
zero_var_cols <- names(numeric_predictors)[sapply(numeric_predictors, function(x) {
  v <- var(x, na.rm = TRUE)
  is.na(v) || v == 0
})]
if (length(zero_var_cols) > 0) {
  message("Dropping zero-variance / all-NA columns: ",
          paste(zero_var_cols, collapse = ", "))
  numeric_predictors <- numeric_predictors %>% select(-all_of(zero_var_cols))
}

# Pearson correlation. `pairwise.complete.obs` handles NAs column-by-column,
# which matters for SAT_AVG / ADM_RATE (sparse by construction).
cor_mat <- cor(numeric_predictors,
               use = "pairwise.complete.obs",
               method = "pearson")

# Any residual NAs come from pairs with no overlapping non-NA rows -- drop
# those columns too so clustering can proceed.
bad_rows <- rowSums(is.na(cor_mat)) > 0
if (any(bad_rows)) {
  message("Dropping columns with NA correlations (no overlapping data): ",
          paste(rownames(cor_mat)[bad_rows], collapse = ", "))
  cor_mat <- cor_mat[!bad_rows, !bad_rows]
}

# Hierarchical clustering on (1 - r)/2 distance so correlated blocks sit
# next to each other -- makes visual scanning for multicollinearity easier.
hc <- hclust(as.dist((1 - cor_mat) / 2))
ord <- hc$order
cor_mat <- cor_mat[ord, ord]

cor_df <- as.data.frame(as.table(cor_mat)) %>%
  rename(var1 = Var1, var2 = Var2, r = Freq)

ggplot(cor_df, aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "white", linewidth = 0.15) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Pearson correlation among continuous predictors",
       subtitle = "Hierarchical clustering ordering",
       x = NULL, y = NULL, fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7),
        panel.grid   = element_blank())
```

Pairs with |r| > 0.8 -- candidates to drop / merge before LASSO. Upper-triangle only so each pair shows once.

```{r}
cor_df %>%
  filter(as.character(var1) < as.character(var2), abs(r) > 0.8) %>%
  arrange(desc(abs(r))) %>%
  mutate(r = round(r, 3)) %>%
  print(n = Inf)
```



